---
title: "dplyr for Data Manipulation"
author: "Jason Baik"
date: "August 5, 2017"
output: html_document
---

```{r include=FALSE}
tutorial::go_interactive()
```

# **What is dplyr?** 
* developed by Hadley Wickham of R Studio 
* provides a "grammar for data manipulation and for operating on data frames"
* some key verbs include: 
1. select(): return a subset of columns of a data frame
2. filter(): extract a subset of rows from a data frame
3. arrange(): reorder rows of a data frame
4. mutate(): add new variables / columns or transform existing variables
5. summarise() / summarize(): generate a summary statistics of different variables in the data frame
6. %>%: the "pipe" operator, used to connect multiple verb actions together.

### Installing dplyr

install.packages("dplyr")  
OR    
install.packages("devtools")  
library(devtools)  
install_github("hadley/dplyr")


### select()
* used to "select" columns of a data frame.
* allows you to get few columns you want to focus on.

> Additional Options: a number of special functions that only work inside select()  

1. starts_with() / ends_with(): starts/ends with a prefix
2. contains(): contains a literal string
3. matches(): matches a regular expression
4. one_of(): variables in character vector
5. everything(): all variables
6. %in%: a more intuitive interface as a binary operator, which returns a logical vector indicating if there is a match or not for its left operand.

```{r}
# Let's select the columns in the diamonds datatset that starts with the letter c!!!
library(ggplot2) #load this package to retrieve diamonds dataset
library(dplyr)
diamonds.f <- select(diamonds, starts_with("c"))
names(diamonds.f)
```

```{r}
library(dplyr)
names(iris)[1:3]
iris.1 <- select(iris, Sepal.Length:Petal.Length)
head(iris.1)
```

#### omit columns using the negative sign!

```{r}
library(dplyr)
names(iris)[1:3]
iris.1 <- select(iris, -(Sepal.Length:Petal.Length))
head(iris.1)
```

### filter()
* used to extract rows from a dataframe.
* faster than subset()

```{r}
library(dplyr)
iris.1 <- filter(iris, Species == "setosa")
head(iris.1)
```

### arrange()
* used to reorder rows of a dataframe according to one of the columns.
* can be arranged in increasing and decreasing order.

```{r}
library(dplyr)
iris.1 <- arrange(iris, Sepal.Length)
iris.2 <- arrange(iris, desc(Sepal.Length))
head(iris.1)
head(iris.2)
```

### rename()
* have new name on left-hand side of the = sign and old name on the right-hand side.

```{r}
library(dplyr)
iris <- rename(iris, SepalL = Sepal.Length, SepalW = Sepal.Width, PetalL = Petal.Length, PetalW = Petal.Width)
head(iris)
```

### mutate()
* computes transformation of variables in a data frame

```{r}
library(dplyr)
iris <- mutate(iris, Sepal_avg = mean(Sepal.Length))
head(iris)
```

### transmute(): same as mutate, but drops all the other unchanged variables

```{r}
library(dplyr)
iris <- transmute(iris, Sepal_avg = mean(Sepal.Length))
head(iris)
```

### Pipeline Operator (%>%)
* imported from the magrittr package
* handy for stringing together multiple dplyr functions in a sequence

```{r}
library(dplyr)
iris <- mutate(iris, Sepal_avg = mean(Sepal.Length)) %>% summarise(median_sepal = median(Sepal_avg))
iris
```

### Additional useful functions 

* summarise(): creates summary statistics for a given column in dataframe such as finding the mean
* group_by(): split dataframe by some variable, apply a function & combine output
* sample_n: select random rows from a dataframe.