---
title: "A Quick Introduction to R (content borrowed from R Programming for Data  Science by Roger D. Peng"
author: "Jason Baik (joonwoob@andrew.cmu.edu)" 
date: "July 28, 2017"
output: html_document
---

```{r include=FALSE}
tutorial::go_interactive()
```
## What is R?
* A dialect of S (programming language developed by John Chambers and only available in a commercial package, S-PLUS)
* Created by Ross Ihaka and Robert Gentleman in Department of Statisitcs at University of Auckland.  
* First announcement of R to public in 1993.  

## Assignment Operator Explained
An assignment operator is used to put values into variables. It is written as <- or = (<- is preferred). A shortcut: Alt - (Windows) Option - (Mac)
This puts spaces before and after the assignment operator.


```{r ex="create_a", type="pre-exercise-code"}
b <- 5
```

```{r ex="create_a", type="sample-code"}
# Create a variable a, equal to 5


# Print out a

```

```{r ex="create_a", type="solution"}
# Create a variable a, equal to 5
a <- 5

# Print out a
a
```

```{r ex="create_a", type="sct"}
test_object("a")
test_output_contains("a", incorrect_msg = "Make sure to print `a`.")
success_msg("Great!")
```

## Comments?
* You can write comments in R with the hashtag, #. Anything coded after # is ignored by R
* Shortcut to comment a line is Ctrl-Shift-C (Windows), Cmd+Shift+C (Mac)

## Basic (*atomic*) Objects in R
1. Character
  + "This is a character"
2. Numeric
  + Numbers in R are treated as Numeric objects 
  + Ex) 1 as 1.00 and 2 as 2.00
3. Integer
  + If you want an integer, you need to specify the L suffix. 
  + Ex) 1L is an integer, 1 is NOT. 2L is an integer, 2 is NOT.
4. Complex
  + Ex) 1 + 0i, 1 + 4i
5. Logical (Booleans)
  + TRUE or...FALSE!
  
## Vectors
* Most basic type of R object
* Can be vector of characters, logical, integers, complex, or numeric
* Vector can only contain objects of the same class. IF you want to include different classes, use lists. More on that later in the course.
* Empty vector is created with vector() function

## **Review Exercise**

```{r ex="create_vector", type="pre-exercise-code"}

```

```{r ex="create_vector", type="sample-code"}
# Create a vector of characters of length 2
# Hint: vector("XXX", length = 2)


# Create a vector of integers of length 10
# Hint: no more hints!
```

```{r ex="create_vector", type="solution"}
# Create a vector of characters of length 2.
vector("character", length = 2)

# Create a vector of integers of length 10.
vector("integer", length = 10)
```

```{r ex="create_vector", type="sct"}
test_output_contains(c(vector("character", length = 2), vector("integer", length = 10)),  incorrect_msg = "Make sure to use vector()")
success_msg("Great Job!")
```

## Attributes
* Attributes are metadata for object as they help describe the object 
1. Names
2. Dimensions
3. Class
4. Length

##### *These can be accessed with the function attributes()*

## **Back to Vectors**
Remember, vectors can not contain objects of different classes. When such happens, *implicit coercion* occurs. This is when R tries to represent objects of different classes in a reasonable fashion.

####Examples of implicit coercion
```{r}
# See what happens to one of the elements when you print y. (Hint: Implicit coercion)
y <- c(34, "a")
```

```{r}
# See what happens to one of the elements when you print a. (Hint: Implicit coercion)
a <- c("a", FALSE)
```

## Explicit Coercion  
Why rely on R to implicitly coerce when you can control coercion yourself?

1. as.character(): coerces into a character
2. as.logical(): coerces into a logical
3. as.numeric(): coerces into a numeric
4. as.complex(): coerces into a complex
5. as.integer(): coerces into an integer

# **Review Exercise**

```{r ex="explicit_coercion", type="pre-exercise-code"}
a <- c(1:5)
```

```{r ex="explicit_coercion", type="sample-code"}
# a is pre-defined in the server as a vector of numbers 1 to 5.
# Coerce a into characters

# Coerce a into integers

```

```{r ex="explicit_coercion", type="solution"}
# a is pre-defined in the server as a vector of numbers 1 to 5.
# Coerce a into characters
as.character(a)
# Coerce a into integers
as.integer(a)
```

```{r ex="explicit_coercion", type="sct"}
test_student_typed(c(as.character(a), as.integer(a)), not_typed_msg = "Make sure to use as.character() and as.integer()")
success_msg("Excellent!")
```

## Matrices
* Vectors with a dimension attribute
* Example: 
```{r} 
matrix(1:8, 2, 4)
```

* As you may see above, matrixes are constructed column-wise, meaning that entries are computed starting in upper-left corner and running down columns.
* They can be modified by using cbind() *(column-binding)* and rbind() *(row-binding)* 

```{r}
x <- c(1:3)
y <- c(4:6)
cbind(x, y)
rbind(x, y)
```

## Lists (remember?)
* Vectors that contain elements of different classes. A dataframe is considered a list.
* Created using list()

```{r}
x <- list("Jason", 2, 2L)
```

## Factors
* represent categorical data
* created with factor()
* when you read in data, characters will be treated as factors

## Missing Values
* is.na() is used to test for NA.
* is.nan() is used to test for NAN.

```{r}
# create a vector x with 3 NA values.
x <- c(1,4,NA,NA,NA)

# check if each element in x is NA. Outputs are boolean values.
is.na(x)

# In x, list the elements that are not NA .
x[!is.na(x)]
```

## Data Frames
* Dataframes store tabular data in R
* Hadley Wickham's dplyr package is effective for working with Data Frames
* Created with data.frame()
* Convert into matrix with data.matrix()

```{r}
# let's create dataframe with 1st column name as crosby and 2nd column name as malkin
x <- data.frame(crosby = 1:5,
       malkin = 6:10)
```

## **Names**
* Can assign names to objects
* Dataframes: Column names are set with names() & Row names are set with row.names()
* Matrix: Column names are set with colnames() & Row names are set with rownames()
```{r}
# let's create a dataframe 
x <- data.frame(crosby = 1:5,
       malkin = 6:10)

# manually change row names of the two columns.
row.names(x) <- c("goals", "assists", "points", "plus", "minus")

# manually change column names
names(x) <- c("murray", "fleury")
```

